Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-06-11 paper/paper.tex
finished my pass

\subsection{Markov decision processes}
Model-free RL has had great success in many real-world problems,
including robotics \citep{ng2004,abbeel2004},
super-human performance in complex video games \citep{mnih2015},
and strategic board games \citep{silver2016mastering}.
We emphasize that the brain in general, and the DMN in particular,
is a physical system governed by the laws of
physics and can be formally described
by Markov processes at a sufficiently coarse scale.
It has indeed been previously proposed \citep{tegmark2016improved} that
any system obeying the laws of classical physics can be accurately modeled as a Markov process as long as the time
step is sufficiently short.
% , defining $s(t)$, the state at time $t$, as the position in state space.
The process has $memory$ if the next state depends not only on the current state
but also on a finite number of past states.
Rational probabilistic planning can be reformulated
as a standard memoryless Markov process by simply expanding the
definition of the state $s$ to include experience episodes of the past.

In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Markov decision processes} (MDPs) \citep{sutton1998reinforcement}.
% For the remainder of this paper, for the sake of simplicity
% we use the abbreviation ``MDP'' to refer to the special case
% of \textit{Partially Observable Markov Decision Processes}.
An MDP operationalizes a sequential decision process
in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., is likely to be different for another agent in the same state).
This is performed alongside each transition in the state space of the environment
 and maintains a probability distribution over the set of possible
states.
% , based on a set of observations and observation probabilities.
This is a minimal set of assumptions that can be made about an environment.

Model-free RL can be plausibly realized in the human brain \citep{doherty2015structure}. Indeed,
a core property of human intelligence has been proposed to underlie improvement
of expected utility outcomes as a strategy for action choice in uncertain
environments \citep{gershman2015computational}, a situation perfectly captured by the formalism of MDPs.
It has also long been proposed \citep{dayan2008decision} that there
is a rather direct mapping of model-free RL learning algorithms
onto aspects of the brain.
The neurotransmitter dopamine could serve
as a ``teaching signal'' to better estimate value associations
and action policies by controlling
synaptic plasticity in the reward-processing circuitry, including the NAc.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's environment or constraints on the state space and transitions between states.
% For instance, if a rat in a maze knows
% that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
%

@@

\subsubsection{Accumulated rewards and policies}
The behavior of the agent is governed by a \textit{policy}, which maps states of the world
to probability distributions over actions.
Starting at time $t=0$,
following a policy $\pi$ generates a trajectory of action choices as follows:
\begin{eqnarray*}
  \begin{split}
    \textbf{choose action: }a_0 &\sim \pi(a|s_0)\\
    \textbf{ observe transition: }s_1 &\sim p(s|s_0,a_0)\textbf{ and collect reward }R_0 = r(s_0, a_0, s_1)\\
    \textbf{choose action: }a_1 &\sim \pi(a|s_1)\\
    \textbf{ observe transition: }s_2 &\sim p(s|s_1,a_1), \textbf{ and collect reward }R_1 = r(s_1, a_1,s_2)\\
    \vdots\\
    \textbf{choose action: }a_{t} &\sim \pi(a|s_{t})\\
    \textbf{ observe transition: }s_{t+1} &\sim p(s|s_{t},a_{t}), \textbf{ and collect reward }R_{t} = r(s_{t}, a_{t},s_{t+1})\\
    \vdots
  \end{split}
\end{eqnarray*}
We assume time-invariance
in that we expect the dynamics of the process
to be equivalent over sufficiently long time windows of equal length (i.e., stationarity).

Since an action executed in the present moment might have repercussions in the far future, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
\textit{cumulative reward} estimate which takes into account expected reward
from action choices in the future.
A common approach to
modeling this accumulation is the time-discounted cumulative reward% function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t = R_0 + \gamma R_1 + \gamma^2 R_2 + \ldots + \gamma^tR_t + \ldots
\end{equation}
This random variable\footnote{Random as it depends both on the environment's dynamic and the
  policy $\pi$ being played (which can be stochastic).}  measures the cumulative reward of
following an action policy $\pi$.
%
Note that value buffering may be realized in the vmPFC.
This DMN node has direct connections to
to the NAc, known to be involved in reward evaluation.

@@

% \subsection{Learning in the DMN}
\subsubsection{Value approximation and the policy matrix}
As already mentioned in the previous section, Q-learning optimizes over the class of
deterministic policies of the form \eqref{eq:qlearning}. State spaces may be extremely large,
and tracking all possible states and actions may require excessive
computation and memory resources.
The need of maintaining an explicit table of
states can be eliminated by instead using of an approximate $Q$-value function $\tilde{Q}(s,a|\theta)$
by keeping track of an approximating parameter $\theta$ of much lower dimension than the number of states.
At a given time step, the world is in a state $s \in \mathcal S$, and the agent takes an
action which it expects to be the most valuable on average, namely

\begin{equation}
  \pi^{\text{hard-max}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
This defines a mapping from states directly to actions.
% $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$.
For instance, a simple linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as was previously proposed \citep{songNIPS2016},
or artificial neural-network models as demonstrated in recent
seminal investigations
\citep{mnih2015,silver2016mastering} for playing complex games (atari, Go, etc.) at super-human levels.
%  and Parr et al. 2003 (JMLR)].
In the DMN, the dmPFC would implement such a hard-max lookup
over the action space.
The model
parameters $\theta$ would correspond to synaptic weights and connection strengths between
brain regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the hard-max policy \eqref{eq:policy}.
The approximating $Q$-value function $\tilde{Q}(s, a|\theta)$ would tell the DMN the (expected) usefulness of taking an action $a$ in state $s$.
The DMN, and in particular its dmPFC node, could then contribute to the choice, at a given state $s$, of an action $a$ which maximizes these approximate
$Q$-values.
It is this mapping from states to actions that is conventionally called
policy matrix \citep{mnih2015,silver2016mastering}.
Learning consists in starting from a given table and
updating it during action choices,
which take the agent to different table entries.

@@

  \subsubsection{Self-training and the loss function}
  \label{sec:self}
Successful learning in brains and computer algorithms is not possible without a
defined learning goal \textemdash the \textit{loss function}.
The action $a$ chosen in state $s$ according to the policy matrix defined in
\eqref{eq:policy} yields a reward $r$ collected by the agent,
% (measured on their internal currency of worldly values ?),
after which the environment transitions to a new state $s' \in \mathcal S$.
% \textbf{XXX Please try to smoothen the previous sentence}
One such cycle yields a new \textit{experience} $e = (s,a,r,s')$.
Each cycle represents a behavior unit of the agent
and is recorded in replay memory buffer \textemdash
which we hypothesize to be subserved by the
HC \textemdash, possibly discarding the oldest entries to make space:
$\mathcal D \leftarrow \text{append}(\mathcal D, e)$.

At time step $k$, the agent seeks an update $\theta_{k} \leftarrow \theta_{k-1} + \delta \theta_{k}$ of the parameters for its approximate model of the $Q$-value function. This warrants a learning process
and definition of a loss function.
% \subsubsection{Stochastic gradient in the DMN via experience replay}
The Bellman equation \eqref{eq:bellman} provides a way to obtain such a loss function \eqref{eq:oracle} as we outline in the following.
% A practically successful RL approach
% needs ways to efficiently approximate the state-action value function
% to allow for scaling to large state and action spaces.
Experience replay consists in sampling
% (uniform or importance-weighted
% \footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.})
batches of experiences $e$
$(s, a, r, s') \sim \mathcal D$ from the replay memory $\mathcal D$.
The agent then tries to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by the Bellman equation \eqref{eq:bellman}, namely
\begin{equation}
  y_k := y_k(s,a,s') =  r + \gamma \max_{a'} \tilde{Q}(s', a'|\theta_{k-1}),
\end{equation}
with the prediction of a parametrized regression model $(s,a)
\mapsto \tilde{Q}(s, a|\theta_{k-1})$.
From a neurobiological perspective,
experience replay can be manifested as the re-occurrence of
neuron spiking sequences that have also occurred during specific prior actions
and environmental states.
%, but the replay has a much faster time scale.
The HC is a strong candidate to contribute to such a mechanism
because neuroscience experiments have repeatedly indicated
in rats, mice, cats, rabbits, songbirds, and
monkeys \citep{buhry2011,nokia2010,dave2000,skaggs2007}.

@@

\subsubsection{Optimal control via stochastic gradient descent in the DMN}
Efficient learning of the entire set of model parameters can effectively be achieved
via stochastic \textit{gradient descent}, a common solver algorithm for finding
local minima based on the first derivate of the optimization objective.
Stochastic here means that the true gradient is estimated from batches of training examples samples,
which, in our case, corresponds to blocks of experience from the replay memory:

@@

% \paragraph{Parameter update.}
\begin{equation}
  \delta= -\alpha_k \nabla_{\theta_{k}}\mathcal L(\theta_{k})
  = -\alpha_k\mathbb E_{(s, a, r, s') \sim \mathcal D}[\underbrace{(\tilde{Q}(s, a|\theta_{k}) - y_k)}_{\text{prediction error}}
    \underbrace{\nabla_{\theta_{k}}\tilde{Q}(s, a|\theta_{k})}_{\text{aversion}}],
  \label{eq:oracle}
\end{equation}
where the positive constants $\alpha_1, \alpha_2,\ldots$ are learning rates.
Thus, the next action is taken to drive reward prediction errors
to percolate from lower to higher processing layers to modulate the
choice of future actions. It is a standard result that under special conditions on the learning rates $\alpha_k$
(namely that the sum $\sum_{k=0}^\infty\alpha_k$ diverges while $\sum_{k=0}^\infty\alpha_k^2$ converges) over time (i.e., in the limit $k \rightarrow \infty$)
the thus generated approximating sequence of $Q$-value functions
% \footnote{The proof is standard and relies on the Banach Fixed-Point Theorem applied to the Bellman transform.}
$$\tilde{Q}(.,.|\theta_0) \rightarrow \tilde{Q}(.,.|\theta_1) \rightarrow \tilde{Q}(.,.|\theta_2) \rightarrow \ldots$$
are attracted and absorbed by the optimal $Q$-value function $Q^*$ defined implicitly by the Bellman equation \eqref{eq:bellman}.

@@

Since theories based on the FEP \citep{friston2010free,fristonAIorRL}
conceptualize ongoing behavior
in an organism to be geared towards the surprise-minimizing goal,
an organism entering a dark room
% \footnote{\textit{Dark} is used here as an allegory to indicate the complete absence of any \textit{external} sensory input.}
(Fig. \ref{fig:darkroom}) would strive to remain in this location
because its sensory inputs are perfectly predictable
given the environmental state \citep{darkroom2012}.
However, such a behavioral tendency is
seldom observed in animals or humans in the real world: In a dark room,
\begin{wrapfigure}{r}{.45\linewidth}
  \centering
  \includegraphics[width=1.\linewidth]{darkroom.jpg}
  \caption{\textbf{The dark room experiment.} An intelligent agent situated
  in a light-deprived closed space can be used
  as a thought experiment for the complete absence
  of external sensory input.}
  \label{fig:darkroom}
\end{wrapfigure}
intelligent agents would search for light sources to \textit{explore}
its surroundings or leave it.
Defenders of the FEP have retorted by advancing the ``full package'' \citep{darkroom2012}:
FEP is proposed to be multi-scale and there would be a meta-scale at which the organism would
be \textit{surprised} by such a \textit{lack of surprise}. According to this argument, a dark room would paradoxically correspond to a state of particularly high surprise.
Driven by surprise-minimization objective, the FEP agent would eventually bootstrap itself out of such saddle points to explore more interesting parts of the environment.
% Recourse to this argument is equivalent to invoking a \textit{halting oracle}\citep{haltingoracle}, an arbitrarily powerful machine which can solve the \textit{halting problem} (an arbitrarily difficult computational problem which cannot be solved under any reasonable \textit{model of computation}, not even in principle!). But this is problematic as it sucks up the problem (that of decision-making and active agency) the theory intends to elude in the first place.
In contrast, an organism operating under our RL-based theory would inevitably
identify the sensory-stimulus-deprived room as a local minimum. Indeed, hippocampal experience replay (see \ref{sec:self}) would serve to sample memories or fantasies of alternative situations
with reward structure. Such artificially generated \textit{internal} sensory input
subserved by the DMN can entice the organism to explore the room,
for instance by looking for and using the light switch or simply finding the room exit.

Finally, we note that FEP and active inference can be reframed in terms of
our model-free RL framework.
This becomes possible by recasting the reward signal maximized by the DMN to 
correspond to negative surprise,
that is, the log-likehood of current sensory priors
the agent has about the world. More explicitly, this corresponds to

\begin{equation}
  r^{\text{FEP}} = \underbrace{\log (p_G)}_{-\text{FEP generative surprise}}.
    \end{equation}
% \paragraph{Comparison to our proposed theory.}
% On the surface, a common point between the FEP and our proposed RL-based framework
%   is placing the minimization of a surprise signal at the core of brain function.
%   Indeed in RL, surprise minimization is subsumed  by accurate prediction of
%   rewarding outcomes in the future. A ``free-energy'' agent is barely a biomechanical machine which has the tendency to resist undesired phase-transitions
%   \citep{friston2010free,fristonAIorRL,ortega2013thermodynamics}.
Such a surprise-guided reinforcement learning scheme has previously been advocated
under the equivalent framework of information compression \citep{schmidhuber2010,mohamed2015}.
Nevertheless, minimization of surprise quantities alone
may be insufficient to explain
the diversity of behaviors observed in humans and other intelligent animals.
% The dark-room experiment, for example, provides a formidable argument against in favor of the insufficiency of the FEP and active inference theories to explain intelligent behavior in humans: when a human is put in a dark room, they don't just sit down in a corner forever, what would be the FEP-optimal thing to do; they actively search for the light switch, and start exploring the depths of the room, possibly looking for food or the way out!

>>>>

\subsection{Markov decision processes}
Model-free RL has had great success in many real-world problems,
including robotics \citep{ng2004,abbeel2004},
super-human performance in complex video games \citep{mnih2015},
and strategic board games like the breakthrough results upon recently on the game of Go \citep{silver2016mastering}, considered to be a golden benchmark problem in artificial intelligence.
We emphasize that the brain in general, and the DMN in particular,
is a physical system governed by the laws of
physics and can be formally described
by Markov processes at a sufficiently coarse scale.
It has indeed been previously proposed \citep{tegmark2016improved} that
any system obeying the laws of classical physics can be accurately modeled as a Markov process as long as the time
step is sufficiently short.
% , defining $s(t)$, the state at time $t$, as the position in state space.
The process has $memory$ if the next state depends not only on the current state
but also on a finite number of past states.
Rational probabilistic planning can be reformulated
as a standard memoryless Markov process by simply expanding the
definition of the state $s$ to include experience episodes of the past.

In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Markov decision processes} (MDPs) \citep{sutton1998reinforcement}.
% For the remainder of this paper, for the sake of simplicity
% we use the abbreviation ``MDP'' to refer to the special case
% of \textit{Partially Observable Markov Decision Processes}.
An MDP operationalizes a sequential decision process
in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., is likely to be different for another agent in the same state),
by maintaining probability distributions over actions according to their expected utility.
% This is performed alongside each transition in the state space of the environment
%  and maintains a probability distribution over the set of possible
% states.
% % , based on a set of observations and observation probabilities.
This is a minimal set of assumptions that can be made about an environment faced by an agent engaged in active learning.

Model-free RL can be plausibly realized in the human brain \citep{doherty2015structure}. Indeed,
it has been proposed \citep{gershman2015computational} that a core property of human intelligence  underlie improvement
of expected utility outcomes as a strategy for action choice in uncertain
environments, a situation perfectly captured by the formalism of MDPs.
It has also long been proposed \citep{dayan2008decision} that there
is a rather direct mapping of model-free RL learning algorithms
onto aspects of the brain.
The neurotransmitter dopamine could serve
as a ``teaching signal'' to better estimate value associations
and action policies by controlling
synaptic plasticity in the reward-processing circuitry, including the NAc.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's environment or constraints on the state space and transitions between states.
% For instance, if a rat in a maze knows
% that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
%

@@

\subsubsection{Accumulated rewards and policies}
The behavior of the agent is governed by a \textit{policy}, which maps states of the world
to probability distributions over actions.
Starting at time $t=0$,
following a policy $\pi$ generates a trajectory of action choices as follows:
\begin{eqnarray*}
  \begin{split}
    \textbf{choose action: }a_0 &\sim \pi(a|s_0)\\
    \textbf{ observe transition: }s_1 &\sim p(s|s_0,a_0)\textbf{ and collect reward }R_0 = r(s_0, a_0, s_1)\\
    \textbf{choose action: }a_1 &\sim \pi(a|s_1)\\
    \textbf{ observe transition: }s_2 &\sim p(s|s_1,a_1), \textbf{ and collect reward }R_1 = r(s_1, a_1,s_2)\\
    \vdots\\
    \textbf{choose action: }a_{t} &\sim \pi(a|s_{t})\\
    \textbf{ observe transition: }s_{t+1} &\sim p(s|s_{t},a_{t}), \textbf{ and collect reward }R_{t} = r(s_{t}, a_{t},s_{t+1})\\
    \vdots
  \end{split}
\end{eqnarray*}
We assume time-invariance
in that we expect the dynamics of the process
to be equivalent over sufficiently long time windows of equal length (i.e., stationarity).
Since an action executed in the present moment might have repercussions in the far future, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
\textit{cumulative reward} estimate which takes into account expected reward
from action choices in the future.
A common approach to
modeling this accumulation is the time-discounted cumulative reward% function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t = R_0 + \gamma R_1 + \gamma^2 R_2 + \ldots + \gamma^tR_t + \ldots
\end{equation}
This random variable\footnote{Random as it depends both on the environment's dynamic and the
  policy $\pi$ being played (which can be stochastic).}  measures the cumulative reward of
following an action policy $\pi$.
%
Note that value buffering may be realized in the vmPFC.
This DMN node has direct connections to
to the NAc, known to be involved in reward evaluation.

@@

% \subsection{Learning in the DMN}
\subsubsection{Value approximation and the policy matrix}
As already mentioned in the previous section, Q-learning optimizes over the class of
deterministic policies of the form \eqref{eq:qlearning}. State spaces may be extremely large,
and tracking all possible states and actions may require prohibitively excessive
computation and memory resources.
The need of maintaining an explicit table of
states can be eliminated by instead using of an approximate $Q$-value function $\tilde{Q}(s,a|\theta)$
by keeping track of an approximating parameter $\theta$ of much lower dimension than the number of states.
At a given time step, the world is in a state $s \in \mathcal S$, and the agent takes an
action which it expects to be the most valuable on average, namely

\begin{equation}
  \pi^{\text{hard-max}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
This defines a mapping from states directly to actions.
% $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$.
For instance, a simple linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as was previously proposed \citep{songNIPS2016},
or artificial neural-network models as demonstrated in recent
seminal investigations
\citep{mnih2015,silver2016mastering} for playing complex games (atari, Go, etc.) at super-human levels.
%  and Parr et al. 2003 (JMLR)].
In the DMN, the dmPFC would implement such a hard-max lookup
over the action space.
The model
parameters $\theta$ would correspond to synaptic weights and connection strengths within and between
brain regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the hard-max policy \eqref{eq:policy}.
The approximating $Q$-value function $\tilde{Q}(s, a|\theta)$ would tell the DMN the (expected) usefulness of taking an action $a$ in state $s$.
The DMN, and in particular its dmPFC node, could then contribute to the choice, at a given state $s$, of an action $a$ which maximizes these approximate
$Q$-values.
This mapping from states to actions that is conventionally called
\textit{policy matrix} \citep{mnih2015,silver2016mastering}.
Learning consists in starting from a given table and
updating it during action choices,
which take the agent to different table entries.

@@

  \subsubsection{Self-training and the loss function}
  \label{sec:self}
Successful learning in brains and computer algorithms is not possible without a
defined learning goal \textemdash the \textit{loss function}.
The action $a$ chosen in state $s$ according to the policy matrix defined in
\eqref{eq:policy} yields a reward $r$ collected by the agent,
% (measured on their internal currency of worldly values ?),
after which the environment transitions to a new state $s' \in \mathcal S$.
% \textbf{XXX Please try to smoothen the previous sentence}
One such cycle yields a new \textit{experience} $e = (s,a,r,s')$.
Each cycle represents a behavior unit of the agent
and is recorded in replay memory buffer \textemdash
which we hypothesize to be subserved by the
HC \textemdash, possibly discarding the oldest entries to make space:
$\mathcal D \leftarrow \text{append}(\mathcal D, e)$.
At time step $k$, the agent seeks an update $\theta_{k} \leftarrow \theta_{k-1} + \delta \theta_{k}$ of the parameters for its approximate model of the $Q$-value function. This warrants a learning process
and definition of a loss function.
% \subsubsection{Stochastic gradient in the DMN via experience replay}
The Bellman equation \eqref{eq:bellman} provides a way to obtain such a loss function \eqref{eq:oracle} as we outline in the following.
% A practically successful RL approach
% needs ways to efficiently approximate the state-action value function
% to allow for scaling to large state and action spaces.
Experience replay consists in sampling
% (uniform or importance-weighted
% \footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.})
batches of experiences $e$
$(s, a, r, s') \sim \mathcal D$ from the replay memory $\mathcal D$.
The agent then tries to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by the Bellman equation \eqref{eq:bellman}, namely
\begin{equation}
  y_k := y_k(s,a,s') =  r + \gamma \max_{a'} \tilde{Q}(s', a'|\theta_{k-1}),
\end{equation}
with the prediction of a parametrized regression model $(s,a)
\mapsto \tilde{Q}(s, a|\theta_{k-1})$.
From a neurobiological perspective,
experience replay can be manifested as the re-occurrence of
neuron spiking sequences that have also occurred during specific prior actions
and environmental states.
%, but the replay has a much faster time scale.
The HC is a strong candidate to contribute to such a mechanism
because neuroscience experiments have repeatedly indicated
in rats, mice, cats, rabbits, songbirds, and
monkeys \citep{buhry2011,nokia2010,dave2000,skaggs2007}.

@@

\subsubsection{Optimal control via stochastic gradient descent in the DMN}
Efficient learning of the entire set of model parameters can effectively be achieved
via stochastic \textit{gradient descent}, a universal algorithm for finding
local minima based on the first derivative of the optimization objective.
Stochastic here means that the true gradient is estimated from batches of training examples samples,
which, in our case, corresponds to blocks of experience from the replay memory:

@@

% \paragraph{Parameter update.}
\begin{equation}
  \delta= -\alpha_k \nabla_{\theta_{k}}\mathcal L(\theta_{k})
  = -\alpha_k\mathbb E_{(s, a, r, s') \sim \mathcal D}[\underbrace{(\tilde{Q}(s, a|\theta_{k}) - y_k)}_{\text{prediction error}}
    \underbrace{\nabla_{\theta_{k}}\tilde{Q}(s, a|\theta_{k})}_{\text{aversion}}],
  \label{eq:oracle}
\end{equation}
where the positive constants $\alpha_1, \alpha_2,\ldots$ are learning rates.
Thus, the next action is taken to drive reward prediction errors
to percolate from lower to higher processing layers to modulate the
choice of future actions. It is a standard result that under special conditions on the learning rates $\alpha_k$ --namely that the learning rates are neither too large nor too small, or more precisely that the sum $\sum_{k=0}^\infty\alpha_k$ diverges while $\sum_{k=0}^\infty\alpha_k^2$--
the thus generated approximating sequence of $Q$-value functions
% \footnote{The proof is standard and relies on the Banach Fixed-Point Theorem applied to the Bellman transform.}
$$\tilde{Q}(.,.|\theta_0) \rightarrow \tilde{Q}(.,.|\theta_1) \rightarrow \tilde{Q}(.,.|\theta_2) \rightarrow \ldots$$
are attracted and absorbed by the optimal $Q$-value function $Q^*$ defined implicitly by the Bellman equation \eqref{eq:bellman}.

@@

On another front, since theories based on the FEP \citep{friston2010free,fristonAIorRL}
conceptualize ongoing behavior
in an organism to be geared towards the surprise-minimizing goal,
an organism entering a dark room
% \footnote{\textit{Dark} is used here as an allegory to indicate the complete absence of any \textit{external} sensory input.}
(Fig. \ref{fig:darkroom}) would strive to remain in this location
because its sensory inputs are perfectly predictable
given the environmental state \citep{darkroom2012}.
However, such a behavioral tendency is
seldom observed in animals or humans in the real world: In a dark room,
intelligent agents would search for light sources to \textit{explore}
its surroundings or leave it.
Defenders of the FEP have retorted by advancing the ``full package'' \citep{darkroom2012}:
FEP is proposed to be multi-scale and there would be a meta-scale at which the organism would
be \textit{surprised} by such a \textit{lack of surprise}. According to this argument, a dark room would paradoxically correspond to a state of particularly high surprise.
\begin{wrapfigure}{r}{.4\linewidth}
  \caption{\textbf{The dark room experiment.} An intelligent agent situated
  in a light-deprived closed space can be used
  as a thought experiment for the complete absence
  of external sensory input.}
  \label{fig:darkroom}  
  \centering
  \includegraphics[width=1.\linewidth]{darkroom_resized.png}
\end{wrapfigure}
Driven by surprise-minimization objective, the FEP agent would eventually bootstrap itself out of such saddle points to explore more interesting parts of the environment.
% Recourse to this argument is equivalent to invoking a \textit{halting oracle}\citep{haltingoracle}, an arbitrarily powerful machine which can solve the \textit{halting problem} (an arbitrarily difficult computational problem which cannot be solved under any reasonable \textit{model of computation}, not even in principle!). But this is problematic as it sucks up the problem (that of decision-making and active agency) the theory intends to elude in the first place.
In contrast, an organism operating under our RL-based theory would inevitably
identify the sensory-stimulus-deprived room as a local minimum. Indeed, hippocampal experience replay (see \ref{sec:self}) would serve to sample memories or fantasies of alternative situations
with reward structure. Such artificially generated \textit{internal} sensory input
subserved by the DMN can entice the organism to explore the room,
for instance by looking for and using the light switch or simply finding the room exit.

Finally, we note that FEP and active inference can be reframed in terms of
our model-free RL framework.
This becomes possible by recasting the Q-value function (i.e expected long-term reward) maximized by the DMN to 
correspond to negative surprise,
that is, the log-likehood of current sensory priors
the agent has about the world. More explicitly, this corresponds to

\begin{eqnarray*}
  -Q^{\text{FEP}} \sim \underbrace{\log (p_G)}_{\text{FEP generative surprise}}.
    \end{eqnarray*}
% \paragraph{Comparison to our proposed theory.}
% On the surface, a common point between the FEP and our proposed RL-based framework
%   is placing the minimization of a surprise signal at the core of brain function.
%   Indeed in RL, surprise minimization is subsumed  by accurate prediction of
%   rewarding outcomes in the future. A ``free-energy'' agent is barely a biomechanical machine which has the tendency to resist undesired phase-transitions
%   \citep{friston2010free,fristonAIorRL,ortega2013thermodynamics}.
Such a surprise-guided reinforcement learning scheme has previously been advocated
under the equivalent framework of information compression \citep{schmidhuber2010,mohamed2015}.
Nevertheless, minimization of surprise quantities alone
may be insufficient to explain the diversity of behaviors observed in humans and other intelligent animals.
% The dark-room experiment, for example, provides a formidable argument against in favor of the insufficiency of the FEP and active inference theories to explain intelligent behavior in humans: when a human is put in a dark room, they don't just sit down in a corner forever, what would be the FEP-optimal thing to do; they actively search for the light switch, and start exploring the depths of the room, possibly looking for food or the way out!
